% ─── Chapter 2: Game Environment & Battle Mechanics ────────────────
\chapter{Game Environment \& Battle Mechanics}
\label{ch:game_env}

\section{Pok\'{e}mon Battles---Overview}

Pok\'{e}mon battles are turn-based strategy matches.  Each player brings a team
of up to six Pok\'{e}mon, but usually only one is active at a time (singles
format).  The objective is to reduce all opposing Pok\'{e}mon to zero HP
(fainted).

\subsection{Turn Structure}

At the start of each turn, both players simultaneously choose a hidden action:

\begin{itemize}
  \item \textbf{Attack:} Use one of the active Pok\'{e}mon's four available
        moves.
  \item \textbf{Switch:} Replace the active Pok\'{e}mon with another from the
        bench.  Switching happens instantly before any moves are executed.
\end{itemize}

If both players choose to attack, execution order is determined by:

\begin{enumerate}
  \item \textbf{Move priority:} Certain moves (e.g.\ Quick Attack, Protect)
        always go first regardless of speed.
  \item \textbf{Speed stat:} Among equal-priority moves, the faster
        Pok\'{e}mon acts first.
\end{enumerate}

\subsection{Pok\'{e}mon Anatomy}

Each Pok\'{e}mon possesses:

\begin{center}
\small
\begin{tabularx}{\textwidth}{l X}
  \toprule
  \textbf{Attribute} & \textbf{Description} \\
  \midrule
  Types (1--2)        & Determine offensive strengths/weaknesses and
                        defensive resistances. \\
  Four moves          & Damaging attacks or status-altering moves with
                        varying power, accuracy, and effects. \\
  Ability             & A passive effect that influences the battle
                        (e.g.\ Intimidate lowers the opponent's Attack on
                        entry). \\
  Held item           & A consumable or passive item that boosts stats,
                        heals, or triggers effects. \\
  Base stats (6)      & HP, Attack, Defense, Sp.\ Atk, Sp.\ Def, Speed ---
                        determine overall power and role. \\
  \bottomrule
\end{tabularx}
\end{center}

\subsection{Damage \& Typing}

Damage is calculated from Attack vs.\ Defence stats, modified by type
interactions:

\begin{center}
\small
\begin{tabularx}{0.8\textwidth}{l l}
  \toprule
  \textbf{Modifier} & \textbf{Multiplier} \\
  \midrule
  STAB (Same-Type Attack Bonus) & $\times 1.5$ \\
  Super Effective               & $\times 2.0$ \\
  Resisted                      & $\times 0.5$ \\
  Immune                        & $\times 0.0$ \\
  \bottomrule
\end{tabularx}
\end{center}

These modifiers stack multiplicatively, so a doubly super-effective STAB move
deals $1.5 \times 2.0 \times 2.0 = 6\times$ base damage.  Understanding type
matchups is therefore central to competent play.

% TODO: Insert type matchup chart image here.
% Capture from: https://www.typecalendar.com/pokemon-type-chart.html
% or generate a standard 18x18 type chart as a figure.

\subsection{Winning \& Losing}

When a Pok\'{e}mon's HP reaches zero, it faints.  The owning player must send
out their next available Pok\'{e}mon.  The battle ends when one side has no
remaining conscious Pok\'{e}mon.  In the Nuzlocke variant, fainted Pok\'{e}mon
are permanently unavailable for subsequent battles.

\section{Pok\'{e}mon Showdown Simulator}

The project does not interact with an actual game client.  Instead, it uses
\textbf{Pok\'{e}mon Showdown}---an open-source battle simulator that
implements official Pok\'{e}mon battle mechanics with high fidelity.  The
simulator runs as a local Node.js server, enabling fast, programmatic battle
execution.

Key properties:

\begin{itemize}
  \item Supports multiple generations (the project uses \textbf{Gen\,8} /
        \texttt{gen8randombattle} format).
  \item Runs as a WebSocket server, accepting connections from automated
        clients.
  \item Handles all battle mechanics internally: damage rolls, stat stages,
        weather, terrain, status conditions, abilities, and items.
  \item Provides structured protocol messages describing battle state changes
        each turn.
\end{itemize}

For distributed training, \textbf{eight parallel Showdown instances} are spun
up on ports 8000--8007 (see Chapter~\ref{ch:infrastructure}).

\section{poke-env Python Library}

\texttt{poke-env} provides the Python interface to Pok\'{e}mon Showdown:

\begin{itemize}
  \item A structured API for interacting with battles (reading state,
        choosing actions).
  \item Battle state objects that expose team composition, active Pok\'{e}mon,
        moves, HP, status effects, weather, and more.
  \item Built-in player classes including \texttt{RandomPlayer} (uniform
        random over legal orders) and \texttt{SimpleHeuristicsPlayer}
        (rule-based AI considering type effectiveness and switching).
  \item Integration utilities for reinforcement learning environments.
\end{itemize}

The project extends \texttt{poke-env} with a custom player class
(\texttt{RandomNoSwitchPlayer}) that samples uniformly among legal \emph{move}
orders and never voluntarily switches---useful as a baseline opponent that
forces the agent to learn move selection before switching strategy.

\section{Pok\'{e}mon Brilliant Diamond Context}

The target game is Pok\'{e}mon Brilliant Diamond (BDSP), a remake of the
Generation~4 titles.  The choice is motivated by:

\begin{itemize}
  \item A well-documented set of mandatory trainer battles.
  \item Available datasets of trainer teams with exact Pok\'{e}mon, levels,
        movesets, and items (see Chapter~\ref{ch:challenge}).
  \item Gen\,8 mechanics in Showdown provide a close match to BDSP's battle
        system.
\end{itemize}

The gauntlet---the ordered sequence of battles the agent must
complete---comprises 706 trainers extracted from BDSP gameplay data, ordered
by ascending difficulty (level, team size).
