\chapter{Reported results, risks, and roadmap}

\section{Milestone 1 status}
Infrastructure for Milestone~1 is in place (training, logging, curriculum, checkpoints, validation hooks). \texttt{temp.md} still lists gaps for a \textbf{complete gauntlet evaluation protocol}: fixed seeds, canonical agent team definition, and competency gates measured outside noisy training dashboards alone.

\section{Performance narrative (internal notes)}
\texttt{temp.md} and \texttt{temp/ai\_docs/04\_current\_state.md}, \texttt{05\_challenges\_limitations.md} describe a mid-project bottleneck: fast progress against weak random baselines, then a long plateau near $\sim$50\% win rate against stronger mixtures, with high value loss/entropy and low policy loss. Treat those curves as \textbf{experiment-specific}: they predate the current default curriculum and reward scale.

\section{Engineering priorities (synthesized)}
\begin{enumerate}
  \item Harden \textbf{evaluation}: deterministic validation, standardized metrics, BDSP gauntlet runner when ready.
  \item Break \textbf{representation} limits: positional/role embeddings already exist in config; older notes complained about missing PE---verify against the branch you measure.
  \item Improve \textbf{categorical signal}: migration from ad-hoc hashing to JSON-backed vocab (\texttt{src/models/vocab.py}) is underway; document vocab version with each run.
  \item Revisit \textbf{action space}: compress or restructure mega/Z/max slots; surface gimmick availability explicitly.
  \item \textbf{Temporal reasoning}: optional LSTM or alternative memory once connectors are proven stable.
\end{enumerate}

\section{Phased experiment log (\texttt{temp/ai\_docs})}
Short internal phase documents capture baseline runs, validation setup, vocab IDs, logging, embedding audits, position embeddings, compressed actions, memory experiments, observation-pipeline investigations, attention diagnostics, and pre-norm architecture trials. They are referenced qualitatively here; cite filenames when publishing a result.
