\chapter{Technical deep dive (from \texttt{temp/ai\_docs})}

This chapter summarizes internal Markdown under \texttt{temp/ai\_docs/}. File names are stable references even when \texttt{temp/} is Git-ignored.

\section{Document map (\texttt{temp/ai\_docs})}
\footnotesize
\begin{description}[style=nextline,leftmargin=0cm,itemsep=0.15em]
  \item[Core overview] \texttt{00\_executive\_summary.md}; \texttt{01\_architecture\_overview.md}; \texttt{02\_model\_architecture.md}; \texttt{03\_training\_pipeline.md}.
  \item[Status / challenges] \texttt{04\_current\_state.md}; \texttt{05\_challenges\_limitations.md}; \texttt{06\_roadmap.md}; \texttt{07\_presentation\_notes.md}.
  \item[Phased experiments] \texttt{04\_phase1\_baseline.md} through \texttt{11\_phase8\_memory.md} (baseline, validation, vocab, logging, embedding audit, position embeddings, compressed actions, memory).
  \item[Deep dives] \texttt{12\_observation\_pipeline\_investigation.md}; \texttt{13\_attention\_collapse\_analysis.md}; \texttt{14\_prenorm\_2layer\_experiment.md}.
  \item[Config log] \texttt{config\_updates\_and\_analysis.md}.
\end{description}
\normalsize

\section{Observation tensor}
\begin{itemize}
  \item \textbf{obs} --- Float32 array shape $(13,164)$: one global token, six own-team slots, six opponent slots; see \texttt{embedding.py} for exact feature breakdown (HP, stats, types, status, volatiles, boosts, move slots, etc.).
  \item \textbf{species}, \textbf{items}, \textbf{abilities} --- Integer IDs per token from \texttt{EmbeddingVocab} (\texttt{src/models/vocab.py}), padding index 0.
  \item \textbf{action\_mask} --- Length-22 binary mask for Gen-style singles with expanded gimmick actions (exact legality from Showdown/\texttt{poke-env}).
\end{itemize}

\section{Policy trunk}
\texttt{PokemonTransformerModel} concatenates per-token continuous features with three embedding vectors, projects to \texttt{hidden\_dim}, applies a PyTorch \texttt{TransformerEncoder}, takes the CLS/global token representation, optionally runs an LSTM over time when enabled, then feeds policy and value MLP heads. Invalid actions receive large negative logits before sampling.

\section{Curriculum and env bridge}
\texttt{CurriculumSingleAgentWrapper} (\texttt{battle\_env.py}) samples an opponent key per episode according to the stage's \texttt{opponent\_mix}, reuses opponent instances in a pool, and calls \texttt{set\_opponent\_context} so observations encode opponent class. \texttt{env\_bridge.py} applies stage payloads when the \texttt{CurriculumManager} promotes.

\section{Validation artifacts}
JSON manifests under \path{data/validation/} support scheduled checkpoint evaluation (\texttt{Validation\-Schedule\-Config}). See \path{data/validation/README.md} for protocol semantics.

\section{Caveat when reading older \texttt{ai\_docs}}
Some passages still mention Adler-32 hashing, fixed 20\,001 vocabulary, or four transformer layers. The living code may use JSON vocab sizes, default \texttt{num\_transformer\_layers=1}, and different reward scales. Treat formulas and file:line references in Markdown as hints and re-verify in Git.
