% ─── Chapter 11: Reproduction Guide ────────────────────────────────
\chapter{Reproduction Guide}
\label{ch:reproduction}

This chapter provides step-by-step instructions for reproducing the training
environment, running experiments, and evaluating checkpoints.

\section{Prerequisites}

\begin{center}
\small
\begin{tabularx}{0.7\textwidth}{l l}
  \toprule
  \textbf{Requirement} & \textbf{Version / Notes} \\
  \midrule
  Python      & 3.13 (specified in \texttt{.python-version}) \\
  \texttt{uv} & Latest (package manager, replaces pip) \\
  Node.js     & Required for Pok\'{e}mon Showdown \\
  NVIDIA GPU  & Recommended (RTX 3090+ for \texttt{standard} preset) \\
  RAM         & 16\,GB minimum, 64\,GB recommended for 48 envs \\
  \bottomrule
\end{tabularx}
\end{center}

\section{Installation}

\begin{lstlisting}[language=bash]
# Clone the repository
git clone https://github.com/mavgit1/DSPRO2-FS26-Pokemon_RL.git
cd DSPRO2-FS26-Pokemon_RL

# Install dependencies (uses uv, not pip)
uv sync
\end{lstlisting}

\section{Starting the Simulator}

Pok\'{e}mon Showdown servers must be running before training.  The project
provides scripts to manage 8~parallel instances:

\begin{lstlisting}[language=bash]
# Start 8 Showdown servers on ports 8000-8007
./scripts/spin_up_multiple_showdown.sh

# Stop all servers when done
./scripts/kill_all_showdown.sh
\end{lstlisting}

\section{Training Commands}

\begin{lstlisting}[language=bash]
# Quick test run (single env, ~150K steps)
uv run train_battler.py --preset quick

# Standard training (48 envs, 10M steps)
uv run train_battler.py --preset standard

# Full resources (RTX 5090)
uv run train_battler.py --preset optimal

# Reduced RAM usage
uv run train_battler.py --preset memory_safe

# Maximum model size
uv run train_battler.py --preset large
\end{lstlisting}

\section{Resuming from Checkpoint}

Training can be interrupted and resumed without losing progress:

\begin{lstlisting}[language=bash]
# Resume from the latest checkpoint in the same MLflow run
uv run train_battler.py --preset standard \
  --resume-checkpoint latest \
  --mlflow-run-id <RUN_ID>
\end{lstlisting}

The \texttt{<RUN\_ID>} can be found in the MLflow UI.  If omitted, a new run
is created.

\section{Validation}

\begin{lstlisting}[language=bash]
# Validate a specific checkpoint against fixed teams
python scripts/validate_checkpoint.py \
  --checkpoint-path checkpoints/step_XXXX \
  --protocol fixed_paired

# Validate against BDSP trainers
python scripts/validate_bdsp_trainers.py \
  --checkpoint-path checkpoints/step_XXXX
\end{lstlisting}

Validation results are logged to MLflow with per-opponent-type breakdowns.

\section{Linting \& Formatting}

\begin{lstlisting}[language=bash]
uv run ruff check .
uv run ruff format .
\end{lstlisting}

\section{Artifacts}

\begin{center}
\small
\begin{tabularx}{\textwidth}{l X}
  \toprule
  \textbf{Location} & \textbf{Content} \\
  \midrule
  \texttt{checkpoints/}     & Saved model checkpoints and self-play weights
                              (\texttt{selfplay\_latest.pt}).  Gitignored. \\
  \texttt{mlruns/}          & MLflow experiment artifacts and metrics. \\
  \texttt{data/}            & BDSP trainer data, gauntlet order, vocabularies,
                              validation team manifests. \\
  \texttt{data/validation/} & Frozen team JSON files for reproducible
                              evaluation. \\
  \bottomrule
\end{tabularx}
\end{center}

\section{Troubleshooting}

\begin{center}
\small
\begin{tabularx}{\textwidth}{l X}
  \toprule
  \textbf{Issue} & \textbf{Solution} \\
  \midrule
  Showdown connection errors &
    Verify servers are running: \texttt{curl localhost:8000}.  Restart with
    the spin-up script. \\
  Out of memory &
    Reduce \texttt{num\_workers} or use the \texttt{memory\_safe} preset.
    Monitor with \texttt{htop}. \\
  Slow training &
    Check GPU utilisation with \texttt{nvidia-smi}.  Ensure CUDA is
    available and the correct preset is selected. \\
  MLflow not logging &
    Verify \texttt{.env} contains valid MLflow credentials.  Check that
    the MLflow server is reachable. \\
  \bottomrule
\end{tabularx}
\end{center}
